FHJ: A Formal Model for Hierarchical Dispatching and Overriding

### 2018-01-13 ECOOP18/paper.tex
test git
Proof of Determinacy in latex
small fix
triangle->fork

\newcommand\saveSpaceFig{\vspace{-2ex}}

\lstset{ %
	language=Java,                % choose the language of the code
	columns=flexible,
	lineskip=-1pt,
	basicstyle=\ttfamily\small,       % the size of the fonts that are used for the code
	numbers=none,                   % where to put the line-numbers
	numberstyle=\ttfamily\tiny,      % the size of the fonts that are used for the line-numbers
	stepnumber=1,                   % the step between two line-numbers. If it's 1 each line will be numbered
	numbersep=5pt,                  % how far the line-numbers are from the code
	backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
	showspaces=false,               % show spaces adding particular underscores
	showstringspaces=false,         % underline spaces within strings
	showtabs=false,                 % show tabs within strings adding particular underscores
	morekeywords={var,override},
	%  frame=single,                   % adds a frame around the code
	tabsize=2,                  % sets default tabsize to 2 spaces
	captionpos=none,                   % sets the caption-position to bottom
	breaklines=true,                % sets automatic line breaking
	breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
	title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
	escapeinside={(*}{*)},          % if you want to add a comment within your code
	keywordstyle=\sffamily\bfseries,
	aboveskip=3pt,
	belowskip=3pt
	% commentstyle=\color{Gray},
	% stringstyle=\color{Green}
}

@@

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{FHJ: A Formal Model for Hierarchical Dispatching and Overriding}
%\titlerunning{A Sample LIPIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros

\author[1]{John Q. Open}
\author[2]{Joan R. Access}
\affil[1]{Dummy University Computing Laboratory, Address/City, Country\\
  \texttt{open@dummyuniversity.org}}
\affil[2]{Department of Informatics, Dummy College, Address/City, Country\\
  \texttt{access@dummycollege.org}}
\authorrunning{J.\,Q. Open and J.\,R. Access} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{John Q. Open and Joan R. Access}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{Dummy keyword -- please provide 1--5 keywords}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\input{sections/Introduction.tex}
\input{sections/Overview.tex}
\input{sections/Formalization.tex}
\input{sections/Auxiliary.tex}
\input{sections/Discussion.tex}
\input{sections/RelatedWork.tex}
\input{sections/Conclusion.tex}

@@

\bibliography{paper}
\appendix
%\input{sections/Appendix.tex}

\end{document}

>>>>

%\newcommand\saveSpaceFig{\vspace{-2ex}}
\newcommand\saveSpaceFig{\vspace{0ex}}
\newcommand\wordfork{fork}
\newcommand\wordFork{Fork}
\lstset{ %
	language=Java,                % choose the language of the code
	columns=flexible,
	lineskip=-1pt,
	basicstyle=\ttfamily\small,       % the size of the fonts that are used for the code
	numbers=none,                   % where to put the line-numbers
	numberstyle=\ttfamily\tiny,      % the size of the fonts that are used for the line-numbers
	stepnumber=1,                   % the step between two line-numbers. If it's 1 each line will be numbered
	numbersep=5pt,                  % how far the line-numbers are from the code
	backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
	showspaces=false,               % show spaces adding particular underscores
	showstringspaces=false,         % underline spaces within strings
	showtabs=false,                 % show tabs within strings adding particular underscores
	morekeywords={var,override},
	%  frame=single,                   % adds a frame around the code
	tabsize=2,                  % sets default tabsize to 2 spaces
	captionpos=none,                   % sets the caption-position to bottom
	breaklines=true,                % sets automatic line breaking
	breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
	title=\lstname,                 % show the filename of files included with \lstinputlisting; also try caption instead of title
	escapeinside={(*}{*)},          % if you want to add a comment within your code
	keywordstyle=\sffamily\bfseries,
	aboveskip=3pt,
	belowskip=3pt
	% commentstyle=\color{Gray},
	% stringstyle=\color{Green}
}

@@

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{FHJ: A Formal Model for Hierarchical Dispatching and Overriding}
% \titlerunning{A Sample LIPIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

%% Please provide for each author the \author and \affil macro, even when authors have the same affiliation, i.e. for each author there needs to be the  \author and \affil macros
\author[1]{John Q. Open}
\author[2]{Joan R. Access}
\affil[1]{Dummy University Computing Laboratory, Address/City, Country\\
	\texttt{open@dummyuniversity.org}}
\affil[2]{Department of Informatics, Dummy College, Address/City, Country\\
	\texttt{access@dummycollege.org}}
\authorrunning{J.\,Q. Open and J.\,R. Access} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus ‘et. al.’

\Copyright{John Q. Open and Joan R. Access}%mandatory, please use full first names. LIPIcs license is “CC-BY”;  http://creativecommons.org/licenses/by/3.0/

% \subjclass{Dummy classification -- please refer to \url{http://www.acm.org/about/class/ccs98-html}}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as “F.1.1 Models of Computation”.
% \keywords{Dummy keyword -- please provide 1--5 keywords}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\input{sections/Introduction.tex}
\input{sections/Overview.tex}
\input{sections/Formalization.tex}
\input{sections/Auxiliary.tex}
\input{sections/Discussion.tex}
\input{sections/RelatedWork.tex}

\input{sections/Conclusion.tex}

@@

\bibliography{paper}
\appendix
\input{sections/Appendix.tex}
\end{document}

### 2018-01-13 ECOOP18/sections/Conclusion.tex
triangle->fork

\MIM{} is formalized following the style of
Featherweight Java and proved to be sound. A prototype interpreter is
implemented in Scala. We believe that the formalization of
hierarchical dispatching features is general and
can be safely embedded in other OO models, so as to have support for the triangle
inheritance.

>>>>

\MIM{} is formalized following the style of
Featherweight Java and proved to be sound. A prototype interpreter is
implemented in Scala. We believe that the formalization of
hierarchical dispatching features is general and
can be safely embedded in other OO models, so as to have support for the \wordfork{}
inheritance.

### 2018-01-13 ECOOP18/sections/Discussion.tex
triangle->fork

Our formalization has a simple form of abstract methods, which behave similarly to conventional methods with respect to conflicts.
Other languages may behave differently. 
For instance, in Java 8 when putting two identical abstract methods together by multiple inheritance, there is no conflict error. In Figure~\ref{fig:abstractdiamond}, we use italic $\emph{m}$ to denote abstract methods. In both cases, the Java compiler accepts the definition of $C$ and automatically merges the two inherited methods $m$ into a single one. \MIM{} behaves differently from Java in both cases.
In the triangle inheritance case (left), $C$ will have two distinct abstract methods corresponding to $A.m$ and $B.m$. 
In the diamond inheritance case, the definition of $C$ is rejected. 
There are two reasons for this difference in behaviour. Firstly, our formalization just treats abstract methods as concrete methods with an empty body, and that simplifies
the rules and proofs a lot. Secondly, and more importantly, we distinguish and treat differently conflicting methods, since they may represent different operations, even if they are abstract. Thus our model adopts a very conservative behavior rather than automatically merging 
methods by default (as done in many languages). 
Arguably, for the diamond case, as we have mentioned, it is actually an intentional conflict due to the same source $T$.
It is possible to change our model to account for other behaviors for abstract methods, but we view this as a mostly 
orthogonal change to our work, that should not affect the essence of the model presented here.

\begin{figure*}[t]
\saveSpaceFig
    % \nocaptionrule
    % \centering
    \centering
    \begin{minipage}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=1.6cm]{pics/P7.pdf}
    \end{minipage}
    \centering
    \hspace*{2pt}
    \begin{minipage}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=1.4cm]{pics/P8.pdf}
    \end{minipage}  
    \caption{Triangle inheritance (left) and diamond inheritance (right) on abstract methods.}\label{fig:abstractdiamond}
\saveSpaceFig
\end{figure*}

>>>>

Our formalization has a simple form of abstract methods, which behave similarly to conventional methods with respect to conflicts.
Other languages may behave differently. 
For instance, in Java 8 when putting two identical abstract methods together by multiple inheritance, there is no conflict error. In Figure~\ref{fig:abstractdiamond}, we use italic $\emph{m}$ to denote abstract methods. In both cases, the Java compiler accepts the definition of $C$ and automatically merges the two inherited methods $m$ into a single one. \MIM{} behaves differently from Java in both cases.
In the \wordfork{} inheritance case (left), $C$ will have two distinct abstract methods corresponding to $A.m$ and $B.m$. 
In the diamond inheritance case, the definition of $C$ is rejected. 
There are two reasons for this difference in behaviour. Firstly, our formalization just treats abstract methods as concrete methods with an empty body, and that simplifies
the rules and proofs a lot. Secondly, and more importantly, we distinguish and treat differently conflicting methods, since they may represent different operations, even if they are abstract. Thus our model adopts a very conservative behavior rather than automatically merging 
methods by default (as done in many languages). 
Arguably, for the diamond case, as we have mentioned, it is actually an intentional conflict due to the same source $T$.
It is possible to change our model to account for other behaviors for abstract methods, but we view this as a mostly 
orthogonal change to our work, that should not affect the essence of the model presented here.

\begin{figure*}[t]
\saveSpaceFig
    % \nocaptionrule
    % \centering
    \centering
    \begin{minipage}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=1.6cm]{pics/P7.pdf}
    \end{minipage}
    \centering
    \hspace*{2pt}
    \begin{minipage}[t]{0.45\textwidth}
        \centering
        \includegraphics[width=1.4cm]{pics/P8.pdf}
    \end{minipage}  
    \caption{\wordFork{} inheritance (left) and diamond inheritance (right) on abstract methods.}\label{fig:abstractdiamond}
\saveSpaceFig
\end{figure*}

### 2018-01-13 ECOOP18/sections/Formalization.tex
triangle->fork

\textsc{(T-New)} is the typing rule for object creation $\new{I}$. The
auxiliary function $\canInstantiate(I)$ (see definition in Section~\ref{sec:otherdefs}) checks whether an interface $I$ 
can be instantiated or not. Since triangle inheritance accepts conflicting branches to coexist, the check requires that the most specific method is concrete for each method on each branch.

>>>>

\textsc{(T-New)} is the typing rule for object creation $\new{I}$. The
auxiliary function $\canInstantiate(I)$ (see definition in Section~\ref{sec:otherdefs}) checks whether an interface $I$ 
can be instantiated or not. Since \wordfork{} inheritance accepts conflicting branches to coexist, the check requires that the most specific method is concrete for each method on each branch.

### 2018-01-13 ECOOP18/sections/Introduction.tex
triangle->fork

In contrast to diamond inheritance, the second case of ambiguity
is \textit{unintentional method conflicts}~\cite{scharli03traits}. In
this case, conflicting 
methods do not actually refer to the same feature. 
In a nominal system, methods can be designed for different
functionality, but happen to have the same names (and signatures).
A simple example of this situation is two \lstinline{draw} methods that
are inherited from a deck of cards and a drawable widget. 
In such context, the two \lstinline{draw} methods have very different meanings, 
but they happen to share the same name.
%%This issue was proposed by the trait paper, so-called
When inheritance is used to compose these methods, a compilation 
error happens due to conflicts. However, unlike the diamond problem,
the conflicting methods have very different meanings and do not share a
common parent. We call such a case \textit{triangle inheritance}, in
analogy to diamond inheritance.

@@

C++ and C\# do allow for two
unintentionally conflicting methods to coexist in a class. C\# allows
this by interface multiple inheritance and explicit method
implementations. But since C\# is a single inheritance language, 
it is only possible to \emph{implement} multiple interfaces (but not
multiple classes). %We will have more detailed discussion on this in Section~\ref{sec:relatedwork}.
C++ accepts triangle inheritance and
resolves the ambiguity by specifying the expected path by
\emph{upcasts}. However, neither the C\# or C++ approaches allow
such conflicting methods to be further overridden. 
% However, C++ has
% limited support for virtual methods with unintentional conflicts, and
% it will often throw errors when composing them. This is again
% unsatisfactory because virtual methods are pervasive in OOP and used 
% for code reuse and extensibility. A problem with the C++ approach is
% that programmers can only use either static or dynamic dispatching separately, but dealing
% with unintentional method conflicts seems to require a combination of
% both. 
Some other workarounds or approaches include delegation and
renaming/exclusion in the trait model. However, renaming/exclusion 
can break the subtyping relation between a subclass and its parent.
This is not adequate for the class model commonly used in mainstream 
OOP languages, where the subclass is always expected to be a subtype 
of the parent class.

>>>>

In contrast to diamond inheritance, the second case of ambiguity
is \textit{unintentional method conflicts}~\cite{scharli03traits}. In
this case, conflicting 
methods do not actually refer to the same feature. 
In a nominal system, methods can be designed for different
functionality, but happen to have the same names (and signatures).
A simple example of this situation is two \lstinline{draw} methods that
are inherited from a deck of cards and a drawable widget. 
In such context, the two \lstinline{draw} methods have very different meanings, 
but they happen to share the same name.
%%This issue was proposed by the trait paper, so-called
When inheritance is used to compose these methods, a compilation 
error happens due to conflicts. However, unlike the diamond problem,
the conflicting methods have very different meanings and do not share a
common parent. We call such a case \textit{\wordfork{} inheritance}, in
analogy to diamond inheritance.

@@

C++ and C\# do allow for two
unintentionally conflicting methods to coexist in a class. C\# allows
this by interface multiple inheritance and explicit method
implementations. But since C\# is a single inheritance language, 
it is only possible to \emph{implement} multiple interfaces (but not
multiple classes). %We will have more detailed discussion on this in Section~\ref{sec:relatedwork}.
C++ accepts \wordfork{} inheritance and
resolves the ambiguity by specifying the expected path by
\emph{upcasts}. However, neither the C\# or C++ approaches allow
such conflicting methods to be further overridden. 
% However, C++ has
% limited support for virtual methods with unintentional conflicts, and
% it will often throw errors when composing them. This is again
% unsatisfactory because virtual methods are pervasive in OOP and used 
% for code reuse and extensibility. A problem with the C++ approach is
% that programmers can only use either static or dynamic dispatching separately, but dealing
% with unintentional method conflicts seems to require a combination of
% both. 
Some other workarounds or approaches include delegation and
renaming/exclusion in the trait model. However, renaming/exclusion 
can break the subtyping relation between a subclass and its parent.
This is not adequate for the class model commonly used in mainstream 
OOP languages, where the subclass is always expected to be a subtype 
of the parent class.

### 2018-01-13 ECOOP18/sections/Overview.tex
triangle->fork

This example illustrates the basic form of triangle inheritance, where two unintentionally conflicting methods
are accepted by multiple inheritance. Note that C++ supports this feature and also addresses the
ambiguity by upcasts. The code for the above example in C++ is similar.

@@

\subsubsection{Potential solutions/workarounds in existing languages:} Unfortunately in all languages we have known
(including C++), the existing approaches are unsatisfactory.
One direction is to simply avoid this issue, by putting overriding before inheritance. For example, as shown in Figure~\ref{fig:drawablesafedeck}(middle), we define a new component \lstinline|DrawableRect| that extends \lstinline|Drawable|, which simply draws the deck as a rectangle, and modifies the hierarchy:
\begin{lstlisting}
interface DrawableRect extends Drawable {
  void draw() {
    JFrame frame = new JFrame("Canvas");
    frame.setSize(600, 600);
    frame.getContentPane().setBackground(Color.red);
    frame.getContentPane().add(new Square(10,10,100,100)); ...
  }
}
interface DrawableSafeDeck extends DrawableRect, SafeDeck {}
\end{lstlisting}
This workaround seems to work, but there are severe issues:
\begin{itemize}
	\item It changes the hierarchy and existing code, hence breaks the modularity.
	\item Separate overriding is required to come after the triangle
          inheritance, especially when the implementation needs functionality from both parents. In the above code, we have assumed
	that the overriding is unrelated to \lstinline|Deck|. But when the drawing relies on some information of the \lstinline|Deck|
	object, we have to either introduce field methods for delegation or change the signature of \lstinline|draw| to take a parameter.
	Either way introduces unnecessary complexity and affects extensibility.
\end{itemize}

@@

\subsubsection{A peek at the hierarchical dispatching algorithm:} In \MIM{}, triangle inheritance allows several original methods (branches)
to coexist, and hierarchical dispatch first finds the most specific
original method (branch), then it finds the most specific hierarchical
overriding on that branch.

@@

In  Figure~\ref{fig:examplesmbody}, (a) is the base case for unintentional conflicts, namely the triangle inheritance. (b) uses overriding to explicitly merge the conflicting methods. (c) represents hierarchical overriding. (d) and (e) are two base cases of diamond inheritance in \MIM{} and the definition of each $C$ is rejected because $T$ is an ambiguous parent to $C$. (f) gives a common solution to the diamond as in Java or traits, which is to explicitly override $A.m$ and $B.m$ in $C$. In the last three examples, conflicting methods $A.m$ and $B.m$ should be viewed as
intentional conflicts, as they come from the same source $T$.
% Similar to many OO languages, the model also allows \textit{super method invocation} in a method body. The invocation \lstinline|super.T::m()| will ignore all the subtypes of \lstinline|T|, and only look at \lstinline|T| together with its super interfaces. It should behave the same as \lstinline|new T().m()| in principle.

>>>>

This example illustrates the basic form of \wordfork{} inheritance, where two unintentionally conflicting methods
are accepted by multiple inheritance. Note that C++ supports this feature and also addresses the
ambiguity by upcasts. The code for the above example in C++ is similar.

@@

\subsubsection{Potential solutions/workarounds in existing languages:} Unfortunately in all languages we have known
(including C++), the existing approaches are unsatisfactory.
One direction is to simply avoid this issue, by putting overriding before inheritance. For example, as shown in Figure~\ref{fig:drawablesafedeck}(middle), we define a new component \lstinline|DrawableRect| that extends \lstinline|Drawable|, which simply draws the deck as a rectangle, and modifies the hierarchy:
\begin{lstlisting}
interface DrawableRect extends Drawable {
  void draw() {
    JFrame frame = new JFrame("Canvas");
    frame.setSize(600, 600);
    frame.getContentPane().setBackground(Color.red);
    frame.getContentPane().add(new Square(10,10,100,100)); ...
  }
}
interface DrawableSafeDeck extends DrawableRect, SafeDeck {}
\end{lstlisting}
This workaround seems to work, but there are severe issues:
\begin{itemize}
	\item It changes the hierarchy and existing code, hence breaks the modularity.
	\item Separate overriding is required to come after the \wordfork{} 
          inheritance, especially when the implementation needs functionality from both parents. In the above code, we have assumed
	that the overriding is unrelated to \lstinline|Deck|. But when the drawing relies on some information of the \lstinline|Deck|
	object, we have to either introduce field methods for delegation or change the signature of \lstinline|draw| to take a parameter.
	Either way introduces unnecessary complexity and affects extensibility.
\end{itemize}

@@

\subsubsection{A peek at the hierarchical dispatching algorithm:} In \MIM{}, \wordfork{}  inheritance allows several original methods (branches)
to coexist, and hierarchical dispatch first finds the most specific
original method (branch), then it finds the most specific hierarchical
overriding on that branch.

@@

In  Figure~\ref{fig:examplesmbody}, (a) is the base case for unintentional conflicts, namely the \wordfork{} inheritance. (b) uses overriding to explicitly merge the conflicting methods. (c) represents hierarchical overriding. (d) and (e) are two base cases of diamond inheritance in \MIM{} and the definition of each $C$ is rejected because $T$ is an ambiguous parent to $C$. (f) gives a common solution to the diamond as in Java or traits, which is to explicitly override $A.m$ and $B.m$ in $C$. In the last three examples, conflicting methods $A.m$ and $B.m$ should be viewed as
intentional conflicts, as they come from the same source $T$.
% Similar to many OO languages, the model also allows \textit{super method invocation} in a method body. The invocation \lstinline|super.T::m()| will ignore all the subtypes of \lstinline|T|, and only look at \lstinline|T| together with its super interfaces. It should behave the same as \lstinline|new T().m()| in principle.
